---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
## Global options, load packages
```{r}
library(knitr)
opts_chunk$set(echo = TRUE)
opts_chunk$set(cache = TRUE)
library(tidyverse)
library(lubridate)
```


## Loading and preprocessing the data
### choices:
- cache the csv
- change date from factor to POSIXct
```{r}
activity_raw = read.csv('~/Downloads/activity.csv')
activity = activity_raw %>% 
  mutate(date = as.POSIXct(date))
summary(activity)
```


## What is mean total number of steps taken per day?
- Note: all NAs that occur take up entire days, so I'm removing those days from
the data
```{r}
keepRows = complete.cases(activity)
byDay = activity %>% 
  filter(keepRows) %>% 
  group_by(date) %>% 
  summarise(totalSteps = sum(steps))
ggplot(data = byDay, aes(totalSteps)) + geom_histogram() + ggtitle('histogram of steps per day')
print(paste0('the mean daily steps taken was: ',round(mean(byDay$totalSteps), 2)))
print(paste0('the median daily steps taken was: ',round(median(byDay$totalSteps), 2)))
```


## What is the average daily activity pattern?
```{r}
keepRows = complete.cases(activity)
byInterval = activity %>% 
  filter(keepRows) %>% 
  group_by(interval) %>% 
  summarise(averageSteps = mean(steps),
            medianSteps = median(steps))
ggplot(byInterval, aes(interval, averageSteps)) + geom_line() + ggtitle('Daily Average Activity')
mostSteps = which.max(byInterval$averageSteps)
print(paste0('interval ',byInterval$interval[mostSteps],' had the most average steps, with ',round(byInterval$averageSteps[mostSteps]),'.'))
```


## Imputing missing values
- Note: imputing missing values by taking the median steps taken at that time interval
```{r}
print(paste0('there are ', sum(is.na(activity)), ' NA rows in the data set.'))
activity_imputed = activity %>% 
  left_join(byInterval) %>% 
  select(-averageSteps) %>% 
  mutate(steps = ifelse(is.na(steps), medianSteps, steps)) %>% 
  select(-medianSteps)

byDay_imputed = activity_imputed %>% 
  group_by(date) %>% 
  summarise(totalSteps = sum(steps))

ggplot(byDay_imputed, aes(totalSteps)) + geom_histogram() + ggtitle('histogram of daily average steps')

print(paste0('the mean daily steps taken was: ',round(mean(byDay_imputed$totalSteps), 2)))
print(paste0('the median daily steps taken was: ',round(median(byDay_imputed$totalSteps), 2)))
```
### Impact of imputing NAs:
1. Mean and Median both changed slightly.
2. Both summary statistics dropped from their values when I just dropped NAs.
3. However, the mean saw a bigger drop than the median.


## Are there differences in activity patterns between weekdays and weekends?
```{r}
activity_imputed = activity_imputed %>% 
  mutate(weekDay = wday(date, label = TRUE)) %>% 
  mutate(weekPart = ifelse(weekDay == 'Sun'|weekDay == 'Sat','weekend','weekday')) %>% mutate(weekPart = as.factor(weekPart)) %>% 
  select(-weekDay)

byIntervalByWeekPart = activity_imputed %>% 
  group_by(interval, weekPart) %>% 
  summarise(averageSteps = mean(steps))

ggplot(byIntervalByWeekPart,
       aes(interval, averageSteps)) +
  geom_line() +
  facet_wrap(~ weekPart, ncol = 1) +
  ggtitle('Average Steps By Daypart - Weekend vs. Weekday')
```

